Action Groupoids and Fractional Types

### 2016-07-01 pibackground.tex
messing with agda style etc
updated sec 3 a bit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Programming with Equivalences}

@@

Both classes of programs, 1-combinators $c$ and
2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT
sense. The elements $c$ of 1-combinators denote type isomorphisms. The
elements $\alpha$ of 2-combinators denote equivalences between these
type isomorphisms. Using the 1-combinators, it is possible to write
any reversible boolean function and hence encode arbitrary boolean
functions by a technique that goes back to \citet{Toffoli:1980}. The
2-combinators provide a layer of programs that computes semantic
preserving transformations of 1-combinators. As a small example, let
us abbreviate $\ot \oplus \ot$ as $\mathsf{Bool}$ and examine two
possible implementations of boolean negation. The first directly uses
the primitive combinator
$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to swap the
two values of type $\mathsf{Bool}$; the second use three consecutive
copies of $\swapp$ to achieve the same effect:
\[\begin{array}{rcl}
\mathsf{not_1} &=& \swapp \\
\mathsf{not_2} &=& (\swapp \odot \swapp) \odot \swapp 
\end{array}\]
We can write a 2-combinator whose \emph{type} is $\mathsf{not_2}
\isotwo \mathsf{not_1}$:
\[
\transtwo~(\idldl ~\respstwo~ \idisotwo)~\idldl
\]
which not only shows the equivalence of the two implementations of
negation but also shows \emph{how} to transform one to the other.

@@

A similar calculation shows that the combinator $a_1$ is the identity
permutation; the combinators $a_3$ and $a_6$ also swap two of the
three elements leaving the third intact; and the combinators $a_4$ and
$a_5$ rotate the three elements. We therefore have
$\mathit{order}(a_1)=1$,
$\mathit{order}(a_2)=\mathit{order}(a_3)=\mathit{order}(a_6)=2$, and
$\mathit{order}(a_4)=\mathit{order}(a_5)=3$. More generally, we define
the power and order of 1-combinators as follows.

@@

@@

@@

For an information-theoretic perspective on the language $\Pi$, we
think of a type containing $N$ values as an abstract system that
has~$N$ distinguishable states. According to the conventional theory
of information~\cite{Shannon1948}, the amount of information contained
in each state of a system with $N$ distinguishable states is $\log N$
bits of information. For example, the type $\mathsf{Bool}$ can be
thought of as an abstract system with two distinguishable states
labeled $\mathsf{true}$ and $\mathsf{false}$ each containing
$\log 2 = 1$ bit. Similarly, the type $\mathbb{3}$ can be thought of
as an abstract system with three distinguishable states each
containing $\log 3$ bits. The logarithmic map implies that information
contained in a composite state is the sum of the information contained
in its constituents. For example, the type
$\mathsf{Bool} \otimes \mathbb{3}$ can be thought of a composite
system consisting of two independent unrelated subsystems. Each state
of the composite system therefore contains
$\log (2 * 3) = \log 2 + \log 3$ bits which is the sum of the
information contained in each subsystem. Since all the combinators
preserve cardinality, they are also information-preserving.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Programming with Equivalences} 
\label{sec:pi}

@@

\noindent Both classes of programs, 1-combinators $c$ and
2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT
sense. The elements $c$ of 1-combinators denote type isomorphisms. The
elements $\alpha$ of 2-combinators denote equivalences between these
type isomorphisms. Using the 1-combinators, it is possible to write
any reversible boolean function and hence encode arbitrary boolean
functions by a technique that goes back to \citet{Toffoli:1980}. The
2-combinators provide a layer of programs that computes
semantic-preserving transformations of 1-combinators. As a small
example, let us abbreviate $\ot \oplus \ot$ as $\mathsf{Bool}$ and
examine two possible implementations of boolean negation. The first
directly uses the primitive combinator
$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to swap the
two values of type $\mathsf{Bool}$; the second use three consecutive
copies of $\swapp$ to achieve the same effect:
\[\begin{array}{rcl}
\mathsf{not_1} &=& \swapp \\
\mathsf{not_2} &=& (\swapp \odot \swapp) \odot \swapp 
\end{array}\]
We can write a 2-combinator whose \emph{type} is $\mathsf{not_2}
\isotwo \mathsf{not_1}$:
\[
\transtwo~(\idldl ~\respstwo~ \idisotwo)~\idldl
\]
which not only shows the equivalence of the two implementations of
negation but also shows \emph{how} to transform one to the other.

@@

More generally we can iterate 1-combinators to produce different
reversible functions between finite sets, eventually wrapping around
at some number which represents the \emph{order} of the underlying
permutation.

@@

For our example combinators on the type $\mathbb{3}$, simple traces
using the operational semantics show the combinator $a_1$ is the
identity permutation; the combinators $a_3$ and $a_6$ swap two of the
three elements leaving the third intact; and the combinators $a_4$ and
$a_5$ rotate the three elements. We therefore have
$\mathit{order}(a_1)=1$,
$\mathit{order}(a_2)=\mathit{order}(a_3)=\mathit{order}(a_6)=2$, and
$\mathit{order}(a_4)=\mathit{order}(a_5)=3$.

@@

The 2-combinators, being complete equivalences between
1-combinators~\cite{Carette2016}, also capture equivalences regarding
power of combinators and their order.

@@

\paragraph{Intermezzo.} For an information-theoretic perspective on
the language $\Pi$, we think of a type containing $N$ values as an
abstract system that has~$N$ distinguishable states. According to the
conventional theory of information~\cite{Shannon1948}, the amount of
information contained in each state of a system with $N$
distinguishable states is $\log N$ bits of information. For example,
the type $\mathsf{Bool}$ can be thought of as an abstract system with
two distinguishable states labeled $\mathsf{true}$ and
$\mathsf{false}$ each containing $\log 2 = 1$ bit. Similarly, the type
$\mathbb{3}$ can be thought of as an abstract system with three
distinguishable states each containing $\log 3$ bits. The logarithmic
map implies that information contained in a composite state is the sum
of the information contained in its constituents. For example, the
type $\mathsf{Bool} \otimes \mathbb{3}$ can be thought of a composite
system consisting of two independent unrelated subsystems. Each state
of the composite system therefore contains
$\log (2 * 3) = \log 2 + \log 3$ bits which is the sum of the
information contained in each subsystem. Since all the combinators
preserve cardinality, they are also information-preserving.
